---
title: "NPS Public Water System Analysis"
format: html
editor: 
  markdown: 
    wrap: 72
---

```{r}
library(leaflet)
library(dplyr)
library(sf)
library(htmltools)
library(readr)
library(sf)
library(flextable)
library(purrr)
library(ggplot2)
library(plotly)
library(scales) 
library(ggrepel)
library(stringr)

```

# Sustaining the Unseen: Prioritizing Water Infrastructure in America's National Parks

## 🌍 Why This Project Matters

Our National Parks are global treasures, welcoming over 300 million
visitors annually alongside thousands of resident park rangers and
support staff. While visitors marvel at towering redwoods or dramatic
canyons, an invisible, highly complex web of **Public Water Systems
(PWS)** quietly ensures public health, safety, and operational
continuity.

Managing water in these landscapes is uniquely challenging:
**Infrastructure Age & Isolation:** Many park systems operate in remote
areas, far from municipal grids, relying on localized infrastructure
that is difficult to maintain and monitor. **Climate Vulnerability:**
From mountain springs drying up to surface water reservoirs suffering
from algal blooms and wildfire-related sediment runoff, climate change
is actively threatening these vital resources. **Regulatory Oversight:**
Because these systems serve the public, they must strictly comply with
the EPA’s Safe Drinking Water Act (SDWA).

This project bridges the gap between raw federal environmental databases
and actionable asset management. By linking the **National Park Service
Water Supply Database (WSD)** with the **EPA's Safe Drinking Water
Information System (SDWIS)**, we can isolate, visualize, and prioritize
the water systems that are most critical to park operations and public
safety.

------------------------------------------------------------------------

## 🎯 Identifying the "High-Priority" Systems

To effectively allocate resources, we categorize and prioritize these
water systems based on three primary dimensions:

### 1. Scale of Human Impact (Population Served)

Not all water systems are built equal. We utilize an 11-category
population scale to identify systems serving massive visitor hubs vs.
small backcountry stations: **High-Impact Systems (Class 5 to 11):**
Systems serving over 3,300 to upwards of 1 million people (such as
primary hubs in Yosemite, Grand Canyon, or Yellowstone). A single system
failure here represents a massive public health risk and potential
park-wide closure. **Localized Systems (Class 1 to 4):** Systems serving
under 3,300 people, often supporting remote ranger stations or
campgrounds. While lower in raw numbers, these systems frequently lack
backup sources, making them highly vulnerable to localized
contamination.

### 2. Physical Water Source Vulnerability

The source of a water system dictates its environmental risk profile:
**Surface Water (SW) / Groundwater Under Direct Influence (GWUDI):**
Highly sensitive to immediate climate impacts, such as droughts, flash
floods, agricultural runoff, and surface contamination. **Groundwater
(GW):** Generally more stable, but highly susceptible to long-term
geological contaminants (like arsenic) or persistent chemical
infiltration (like PFAS).

### 3. Historical Compliance & Risk (SDWA Violations)

By analyzing historical violations, we can highlight systems with a
track record of compliance issues, particularly those sensitive to
climate stressors, providing park managers with a data-driven "early
warning" system.

------------------------------------------------------------------------

## 🗺️ Interactive System Explorer

Use the map below to explore public water systems across the National
Park network. **Filter** layers using the top-right control to toggle
park boundaries and water system classifications. **Hover** over any
point to view the system name, exact population served, and size
classification. **Click** on any park boundary to identify the official
administrative unit name.

U.S. Environmental Protection Agency’s (EPA) Safe Drinking Water
Information System (SDWIS) data downloaded from:
<https://echo.epa.gov/tools/data-downloads#drinkingwater>
(SDWA_PUB_WATER_SYSTEMS.csv) and joined with WSD using unique Public
Water System Identification numbers.

```{r}
# Load the tabular metadata (if needed for other plots, otherwise loaded here)
wsd_sdwa <- read_csv("data/wsd_sdwa_updated.csv", show_col_types = FALSE)

# 1. Load your local pre-simplified NPS park boundaries
nps_boundaries <- sf::st_read("data/nps_boundaries.geojson", quiet = TRUE) %>% 
  sf::st_transform(crs = 4326) 

# 2. Load the clean spatial data we exported
map_data <- sf::st_read("data/map_data.geojson", quiet = TRUE) %>% 
  sf::st_transform(crs = 4326)

# Define clean labels for the 11 population scales
pop_labels <- c(
  "<=100", "101-500", "501-1,000", "1,001-3,300", "3,301-10,000", 
  "10,001-50,000", "50,001-100,000", "100,001-250,000", 
  "250,001-500,000", "500,001-1,000,000", ">1,000,000"
)

# Define the master map-building function
build_nps_map <- function(data_input) {
# 3. Add back the hover text dynamically using your existing 'classification' column
map_data <- data_input %>%
  mutate(
    # Re-establish the factor levels so the legend stays in perfect numerical order
    POP_CAT_11_CODE = factor(POP_CAT_11_CODE, levels = as.character(1:11)),
    
    # Create the HTML list column on the fly
    hover_text = lapply(paste0(
      "<strong>System Name:</strong> ", PWS_NAME, "<br/>",
      "<strong>Classification:</strong> ", classification, "<br/>",
      "<strong>Population Served:</strong> ", scales::comma(POPULATION_SERVED_COUNT), "<br/>",
      "<strong>Population Range:</strong> ", POP_CAT_11_DESC
    ), HTML)
  )

# 4. Create palette mapping
pop_palette <- colorFactor(
  palette = "YlOrRd", 
  domain = map_data$POP_CAT_11_CODE
)

# Get unique classifications to build the interactive layers dynamically
classifications <- unique(map_data$classification)

# 5. Initialize the leaflet map
m <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldTopoMap) %>% 
  setView(lng = -98.5795, lat = 39.8283, zoom = 4) %>%
  
  # Add National Park Boundaries
  addPolygons(
    data = nps_boundaries,
    fillColor = "#2a9d8f",             
    fillOpacity = 0.15,                
    color = "#1d3557",                 
    weight = 1.2,                      
    popup = ~UNIT_NAME,                
    group = "Park Boundaries"          
  )

# 6. Dynamically add a separate marker layer for each classification
for (cl in classifications) {
  sub_data <- map_data %>% filter(classification == cl)
  
  m <- m %>% addCircleMarkers(
    data = sub_data,
    lng = ~source_longitude, 
    lat = ~source_latitude,
    radius = 6,                      
    color = "#4a4a4a",               
    fillColor = ~pop_palette(POP_CAT_11_CODE), 
    fillOpacity = 0.8,
    stroke = TRUE,
    weight = 0.8,
    label = ~hover_text,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto"
    ),
    group = cl # Assigns this subset of dots to its own toggle group
  )
}

# 7. Add map styling controls and the interactive legend
m %>%
  addLegend(
    pal = pop_palette, 
    values = map_data$POP_CAT_11_CODE, 
    opacity = 0.8, 
    title = "Population Served",
    labFormat = labelFormat(transform = function(x) pop_labels[as.numeric(x)]),
    position = "bottomright"
  ) %>%
  
  # Interactive Layer Toggles
  addLayersControl(
    overlayGroups = c("Park Boundaries", classifications),
    options = layersControlOptions(collapsed = FALSE),
    position = "topright"
  )
}

build_nps_map(map_data)
```

# High-Impact Infrastructure: Top 10 Water Systems by Population Served

The table below highlights the ten largest public water systems
operating within or adjacent to National Park boundaries. Due to the
scale of human impact, these systems represent the highest priority for
asset management, infrastructure investment, and water quality
monitoring. Compare the full network list with the directly managed
subset to see how operational responsibility shifts across major park
units.

```{r}
# Function to generate a formatted Flextable for top N systems
create_top_systems_table <- function(data_input, n = 10) {
  
  # 1. Extract and clean top N systems
  top_systems <- data_input %>%
    filter(!is.na(POPULATION_SERVED_COUNT)) %>%
    distinct(pws_number, .keep_all = TRUE) %>%
    arrange(desc(POPULATION_SERVED_COUNT)) %>%
    slice(1:n) %>%
    select(
      `System ID` = pws_number,
      `System Name` = PWS_NAME,
      `Park Name` = park_name,
      `Population Served` = POPULATION_SERVED_COUNT,
      `Size Class` = POP_CAT_11_DESC
    )

  # 2. Design and return the Flextable
  flextable(top_systems) %>%
    theme_vanilla() %>%
    colformat_double(j = "Population Served", big.mark = ",", digits = 0) %>%
    bg(bg = "#2a9d8f", part = "header") %>%
    color(color = "white", part = "header") %>%
    bold(part = "header") %>%
    align(align = "center", part = "header") %>%
    align(j = c("System ID", "Population Served", "Size Class"), align = "center", part = "body") %>%
    align(j = c("System Name", "Park Name"), align = "left", part = "body") %>%
    autofit()
}

create_top_systems_table(wsd_sdwa)
```

::: {.callout-note appearance="simple"}
### Why Filter by Source Type?

The complete dataset includes every water system located within or
adjacent to park lands. However, a significant portion of these are
**municipal, hauled, or tribally owned systems**.

While these third-party connections are vital for park operations, their
primary maintenance, compliance, and capital investment responsibilities
rest with external utilities. **Directly managed NPS systems**, on the
other hand, represent assets where park staff hold 100% of the
operational risk, maintenance burden, and regulatory oversight under the
Safe Drinking Water Act (SDWA). Isolating these systems allows asset
managers to focus on infrastructure directly funded and maintained by
the agency.
:::

```{r}
# Filter dataset
map_data_nps_only <- map_data %>%
  filter(!str_detect(source_type, regex("municipal|hauled|tribal", ignore_case = TRUE)) | is.na(source_type))

# Render Map 2 (NPS Managed Only)
build_nps_map(map_data_nps_only)
```

### Top 10 NPS Managed Water Systems By Population Served

```{r}
# Filter tabular dataset using explicit stringr namespace
wsd_sdwa_nps_only <- wsd_sdwa %>%
  filter(!stringr::str_detect(source_type, regex("municipal|hauled|tribal", ignore_case = TRUE)) | is.na(source_type))

# Render filtered table
create_top_systems_table(wsd_sdwa_nps_only)
```

# Human Stressors: Integrating 5-Year Visitor Use Metrics

While understanding a water system's static design population (residents
and staff) is essential, it only tells half the story. The true
operational stress on National Park Service (NPS) water infrastructure
is driven by the millions of recreational tourists who visit these parks
every year.

To quantify this human stressor, we integrated public water system data
with official park visitation datasets sourced from the **NPS Integrated
Resource Management Applications (IRMA) Portal**.

Rather than relying on a single year of visitation data, which is highly
susceptible to external anomalies like extreme weather events,
wildfires, road closures, or economic shifts, we designed a more
resilient metric: a **5-year rolling average (2021–2025) of Total
Visitors (\`TV\`)**.

------------------------------------------------------------------------

```{r}
# 1. Direct URL to the official IRMA Visitor Use Statistics "Main_Data.csv"
# (This contains historical visitation data up through the most recent calendar year)
visitation_url <- "https://irma.nps.gov/DataStore/DownloadFile/756959"

# 2. Read the CSV directly from the web
visitor_data_raw <- read_csv(visitation_url, show_col_types = FALSE)

# 3. Process visitation: Monthly Sums -> Annual Totals -> 5-Year Average
park_visitation_5yr <- visitor_data_raw %>%
  # Filter for the last 5 complete years and the "TV" statistic
  filter(
    Year >= 2021 & Year <= 2025,
    Statistic == "TV"
  ) %>%
  
  # Step A: Sum monthly values to get the total annual visitors for each park and year
  group_by(UnitCode, Year) %>%
  summarize(
    Annual_Visits = sum(Value, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  
  # Step B: Average those annual totals over the 5-year window
  group_by(UnitCode) %>%
  summarize(
    Avg_Annual_Visits_5Yr = mean(Annual_Visits, na.rm = TRUE),
    .groups = "drop"
  ) 

# 4. Merge the datasets
wsd_with_5yr_visitors <- wsd_sdwa_nps_only %>%
  left_join(park_visitation_5yr, by = c("park_unit" = "UnitCode"))

# 5. Save your new master dataset
write_csv(wsd_with_5yr_visitors, "data/wsd_sdwa_visitors.csv")
message("Successfully joined 5-year 'TV' averages and saved to data/wsd_sdwa_visitors.csv")
```

```{r}
# 1. Prepare top 15 parks by 5-year average visitation baseline
top_visitation_parks <- wsd_with_5yr_visitors %>%
  distinct(park_name, Avg_Annual_Visits_5Yr) %>%
  filter(!is.na(Avg_Annual_Visits_5Yr), Avg_Annual_Visits_5Yr > 0) %>%
  arrange(desc(Avg_Annual_Visits_5Yr)) %>%
  slice(1:15)

# 2. Build the Baseline Visitation Bar Chart
ggplot(top_visitation_parks, aes(
  x = reorder(park_name, Avg_Annual_Visits_5Yr), 
  y = Avg_Annual_Visits_5Yr
)) +
  geom_col(fill = "#2a9d8f", width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = label_comma(scale = 1/1e6, suffix = "M")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9.5, face = "italic"),
    axis.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank() # Clean horizontal gridlines
  ) +
  labs(
    title = "5-Year Average Visitor Baseline (2021–2025)",
    subtitle = "Resilient multi-year human stressor metric integrated from NPS IRMA data",
    x = NULL,
    y = "Average Annual Visitors (Millions)"
  )
```

## Methodological Refinement: Allocating Park Visitation to Individual Water Systems

When evaluating public health exposure and infrastructure stress across
national parks, relying solely on **total park-level visitation** can
distort risk assessments. Large, multi-use park units, such as Grand
Canyon, Yellowstone, or Yosemite, host millions of annual visitors
distributed across dozens of geographically isolated water systems.

Assigning a park's total visitor count to every water asset within its
boundaries would severely over-penalize small, low-capacity systems
(e.g., remote trailhead well systems or backcountry housing units)
simply because they reside within a high-traffic park. Conversely, it
underrepresents the true relative burden placed on major utility hubs
that serve primary lodge and visitor center complexes.

To resolve this imbalance, we implement a **Weighted Capacity Allocation
Model** to estimate the true **System-Allocated Visitor Footprint**.

------------------------------------------------------------------------

### 🧮 Mathematical Framework

To distribute park-level visitor traffic across individual assets
proportional to their designed supply responsibility, we calculate a
system's capacity ratio relative to the total NPS-managed infrastructure
capacity within that park unit:

1.  <u style="text-decoration-thickness: 1px; text-underline-offset: 3px;">**Park
    Infrastructure Capacity (**$\text{Pop}_{\text{park}}$):</u> We
    aggregate the Safe Drinking Water Act (SDWA) population served
    across all *directly managed* NPS water systems within a single park
    unit:

$$\text{Pop}_{\text{park}} = \sum_{i=1}^{n} \text{Pop}_{\text{system}_i}$$

2.  <u style="text-decoration-thickness: 1px; text-underline-offset: 3px;">**System
    Capacity Share (**$W_{\text{system}}$):</u> We establish the
    proportional capacity weight ($W$) for each system:

$$W_{\text{system}} = \frac{\text{Pop}_{\text{system}}}{\text{Pop}_{\text{park}}}$$

3.  <u style="text-decoration-thickness: 1px; text-underline-offset: 3px;">**Allocated
    Annual Visitors (**$\text{Vis}_{\text{allocated}}$):</u> We apply
    this capacity weighting factor to the park's 5-year average annual
    visitation ($\text{Vis}_{\text{park}}$):

$$\text{Vis}_{\text{allocated}} = W_{\text{system}} \times \text{Vis}_{\text{park}}$$

------------------------------------------------------------------------

### 🎯 Key Assumptions & Operational Context

-   **Design Capacity as a Demand Proxy:** In SDWA reporting,
    `POPULATION_SERVED_COUNT` reflects peak design capacity or overnight
    service bounds. Using this value to weight visitor demand assumes
    that infrastructure investment naturally scales with human foot
    traffic across park sub-districts.
-   **Granular Risk Scoring:** By shifting from raw park totals to
    `Vis_allocated`, our **System Priority Index (SPI)** accurately
    scores **Systemic Pressure** at the individual asset level. This
    ensures capital improvement recommendations target systems
    experiencing genuine high-volume operational fatigue

```{r}
system_visitor_allocation <- wsd_with_5yr_visitors %>%
  # Ensure population and visitation numbers are numeric and non-NA
  mutate(
    POPULATION_SERVED_COUNT = coalesce(POPULATION_SERVED_COUNT, 0),
    Avg_Annual_Visits_5Yr   = coalesce(Avg_Annual_Visits_5Yr, 0)
  ) %>%
  
  # Group by Park to calculate total park-level infrastructure capacity
  group_by(park_name) %>%
  mutate(
    park_total_pop_served = sum(POPULATION_SERVED_COUNT, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  
  # Calculate System Ratio & Allocated Visitor Footprint
  mutate(
    # Ratio of system capacity to overall park water capacity
    # (Avoid division by zero if a park has 0 recorded population)
    system_pop_ratio = if_else(
      park_total_pop_served > 0, 
      POPULATION_SERVED_COUNT / park_total_pop_served, 
      0
    ),
    
    # Fraction of park visitors allocated to this specific system
    system_allocated_visitors = round(system_pop_ratio * Avg_Annual_Visits_5Yr, 0)
  )
# 2. Extract and format the Top 10 systems by allocated visitors
top_10_allocated <- system_visitor_allocation %>%
  filter(!is.na(pws_number)) %>%
  distinct(pws_number, .keep_all = TRUE) %>%
  arrange(desc(system_allocated_visitors)) %>%
  slice(1:10) %>%
  mutate(
    # Multiply by 100 so 0.25 becomes 25.0
    system_pop_ratio = system_pop_ratio * 100
  ) %>%
  select(
    `System ID`                = pws_number,
    `System Name`              = PWS_NAME,
    `Park Name`                = park_name,
    `System Capacity Share`    = system_pop_ratio,
    `Total Park Visitors`      = Avg_Annual_Visits_5Yr,
    `Allocated Visitor Load`   = system_allocated_visitors
  )
# 3. Design the Flextable
ft_allocated <- flextable(top_10_allocated) %>%
  theme_vanilla() %>%
  
  # Format percentages (now scaled) and big numbers cleanly
  colformat_double(j = "System Capacity Share", digits = 1, suffix = "%") %>%
  colformat_double(j = c("Total Park Visitors", "Allocated Visitor Load"), big.mark = ",", digits = 0) %>%
  
  # Header Styling
  bg(bg = "#2a9d8f", part = "header") %>%
  color(color = "white", part = "header") %>%
  bold(part = "header") %>%
  align(align = "center", part = "header") %>%
  
  # Body Alignment
  align(j = c("System ID", "System Capacity Share", "Total Park Visitors", "Allocated Visitor Load"), align = "center", part = "body") %>%
  align(j = c("System Name", "Park Name"), align = "left", part = "body") %>%
  
  # Highlight the key output column with a light background
  bg(j = "Allocated Visitor Load", bg = "#f4f9f8", part = "body") %>%
  bold(j = "Allocated Visitor Load", part = "body") %>%
  
  autofit()

# Render table
ft_allocated
```

**Reading the System Capacity Share Plot:**

-   **Bar Length (Total Park Load):** The total length of each
    horizontal bar represents the park’s overall 5-year average annual
    visitation.

-   **Color Segments (Individual Systems):** Each distinct colored
    segment within a bar corresponds to an individual public water
    system operating within that park unit. The width of a segment
    reflects its **capacity share ratio**—the proportion of total park
    water capacity managed by that specific system.

    -   **Solid, single-color bars** indicate parks reliant on a
        **centralized water system** carrying 100% of the operational
        footprint (high single-point failure risk).

    -   **Multi-segmented bars** indicate a **decentralized network** of
        smaller systems sharing total park visitor demand (distributed
        operational risk).

-   **Interactive Exploration:** Hover over any colored segment to
    inspect the exact system name, its percentage capacity share, and
    its calculated **Allocated Annual Visitor Load**.

```{r}
# 1. Extract the unique park names directly from your Top 10 table
# Note: Uses `Park Name` if extracted after select(), or `park_name` if extracted before select()
target_parks <- unique(top_10_allocated$`Park Name`)

# 2. Filter system_visitor_allocation for systems within those target parks
plot_data <- system_visitor_allocation %>%
  filter(park_name %in% target_parks) %>%
  mutate(
    # Fallback hierarchy: PWS_NAME -> water_system_name -> pws_number -> Unregistered
    clean_system_name = coalesce(PWS_NAME, water_system_name, pws_number, "Unregistered System")
  )

# 3. Build plot matching the top 10 table
top_parks_plot <- ggplot(plot_data, aes(
  x = reorder(park_name, Avg_Annual_Visits_5Yr), 
  y = system_allocated_visitors, 
  fill = clean_system_name,
  text = paste0("<b>Park:</b> ", park_name, "<br>",
                "<b>System:</b> ", clean_system_name, "<br>",
                "<b>Share:</b> ", round(system_pop_ratio * 100, 1), "%<br>",
                "<b>Allocated Visitors:</b> ", scales::comma(system_allocated_visitors))
)) +
  geom_col(position = "stack", color = "white", linewidth = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "System Capacity Share Across Top 10 Parks",
    x = NULL,
    y = "Allocated Annual Visitors"
  )

# 4. Render interactive plot
ggplotly(top_parks_plot, tooltip = "text") 
```

### Priority Risk Framework: System Priority Index (SPI)

To move beyond isolated metrics, we synthesize structural system
features with human foot traffic into a unified **System Priority Index
(SPI)**. The SPI evaluates each directly managed water asset across
three foundational risk dimensions yielding a scaled composite score
from **0 to 100**::

[1.]{.underline} **Infrastructural Load (35% Weight)**: Local population
capacity served (\`POPULATION_SERVED_COUNT\`). [2.]{.underline}
**Systemic Pressure (35% Weight)**: The proportional 5-year visitor
footprint (\`system_allocated_visitors\`). [3.]{.underline}
**Environmental Exposure (30% Weight)**: Physical water source
vulnerability, penalizing surface water systems due to direct climate
susceptibility (e.g., drought sensitivity, algae blooms, and runoff
contamination) compared to protected groundwater aquifers.

$\text{SPI} = \left( 0.35 \cdot S_{\text{load}} + 0.35 \cdot S_{\text{pressure}} + 0.30 \cdot S_{\text{source}} \right) \times 20$

#### Risk Categorization

-   **Tier 1: Critical Priority (**$\text{SPI} \ge 80$): High-capacity
    surface water or highly stressed groundwater systems requiring
    prioritized capital investment and proactive water quality
    monitoring.

-   **Tier 2: High Priority (**$60 \le \text{SPI} < 80$): Systems
    experiencing substantial visitor load or moderate environmental
    exposure.

-   **Tier 3: Moderate Priority (**$40 \le \text{SPI} < 60$):
    Lower-capacity systems operating under typical baseline conditions.

-   **Tier 4: Low Priority (**$\text{SPI} < 40$): Small groundwater
    systems or third-party purchased connections with low visitor
    demand.

```{r}
spi_dataset <- system_visitor_allocation %>%
  mutate(
    # 1. Map fine-grained source_type into standardized Environmental Risk Categories
    source_category = case_when(
      source_type %in% c("river/stream", "lake/reservoir", "ocean", "rainwater") ~ "Surface / Direct Surface Water",
      source_type %in% c("spring", "reclaimed")                                  ~ "Spring / Reclaimed Water",
      source_type %in% c("deep groundwater", "shallow groundwater", 
                         "unknown groundwater") | is.na(source_type)            ~ "Groundwater (All Depths)",
      source_type == "purchased/concessioner"                                   ~ "Purchased / Concessioner",
      TRUE                                                                      ~ "Groundwater (All Depths)"
    ),
    
    # 2. Assign Granular Environmental Risk Scores (1 to 5)
    score_source = case_when(
      # Highest vulnerability: Direct exposure to weather, runoff, and drought
      source_type %in% c("river/stream", "lake/reservoir", "ocean") ~ 5,
      source_type == "rainwater"                                   ~ 4.5,
      
      # Moderate-High vulnerability: Highly variable flow / complex treatment
      source_type == "spring"                                      ~ 4,
      source_type == "reclaimed"                                   ~ 3.5,
      
      # Moderate-Low vulnerability: Soil-filtered aquifers
      source_type == "shallow groundwater"                         ~ 3,
      source_type %in% c("deep groundwater", "unknown groundwater") ~ 2,
      
      # Lowest direct NPS environmental exposure (third-party delivery)
      source_type == "purchased/concessioner"                      ~ 1,
      
      # Fallback for NA
      TRUE                                                         ~ 2
    ),
    
    # Dimension 2: Local Capacity Score (Quintile Bins 1-5)
    score_load = ntile(POPULATION_SERVED_COUNT, 5),
    
    # Dimension 3: Systemic Pressure Score (Quintile Bins 1-5)
    score_pressure = ntile(system_allocated_visitors, 5),
    
    # Calculate Composite SPI Score (0 - 100 Scale)
    spi_score = round(((0.35 * score_load) + (0.35 * score_pressure) + (0.30 * score_source)) * 20, 1),
    
    # Assign Priority Risk Tiers
    priority_tier = case_when(
      spi_score >= 80 ~ "Tier 1: Critical Priority",
      spi_score >= 60 ~ "Tier 2: High Priority",
      spi_score >= 40 ~ "Tier 3: Moderate Priority",
      TRUE            ~ "Tier 4: Low Priority"
    ),
    
    # Clean system display name
    clean_system_name = coalesce(PWS_NAME, water_system_name, pws_number, "Unregistered System")
  )

# 2. Build Interactive Priority Plot
spi_plot <- ggplot(spi_dataset, aes(
  x = system_allocated_visitors,
  y = spi_score,
  color = source_category,
  size = POPULATION_SERVED_COUNT,
  text = paste0(
    "<b>System:</b> ", clean_system_name, "<br>",
    "<b>Park:</b> ", park_name, "<br>",
    "<b>SPI Score:</b> ", spi_score, " (", priority_tier, ")<br>",
    "<b>Source Type:</b> ", source_category, "<br>",
    "<b>Allocated Visitors:</b> ", scales::comma(system_allocated_visitors), "<br>",
    "<b>Population Served:</b> ", scales::comma(POPULATION_SERVED_COUNT)
  )
)) +
  # Add risk tier threshold reference line
  geom_hline(yintercept = 80, linetype = "dashed", color = "#e76f51", alpha = 0.7) +
  geom_point(alpha = 0.75) +
  scale_x_log10(labels = scales::label_comma()) + # Log-scale x-axis to separate clusters
  scale_color_manual(values = c(
  "Surface / Direct Surface Water" = "#d90429", # Dark Red (High Risk)
  "Spring / Reclaimed Water"       = "#f77f00", # Orange (Moderate-High Risk)
  "Groundwater (All Depths)"       = "#2a9d8f", # Teal (Low Risk)
  "Purchased / Concessioner"       = "#e9c46a"  # Yellow (External)
)) +
  scale_size_continuous(range = c(2, 9), guide = "none") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold")
  ) +
  labs(
    title = "System Priority Index (SPI) vs. Visitor Footprint",
    x = "Allocated Visitor Footprint (Log Scale)",
    y = "System Priority Index Score (0-100)",
    color = "Water Source Type"
  )

# Render interactive Plotly graphic
ggplotly(spi_plot, tooltip = "text")
```
